Practical Optional Types for Clojure

### 2015-02-26 overview.tex
review first half of formal model
better examples. explain syntax

We now begin a tour of the central features of Typed Clojure,
beginning with Clojure itself. In our presentation, we will make 
use of the full Typed Clojure system to illustrate the key type system
ideas, before studying the key features in detail in
section~\ref{sec:formal}.

@@

\citet{TF10}
presented Typed Racket with occurrence typing,
an inference technique for conditional control flow.
They introduce the concept of occurrence typing 
with the following example.

@@

\begin{exmp}
\inputminted[firstline=1]{clojure}{code/demo/src/demo/eg1.clj}
\end{exmp}

Clojure uses \emph{namespaces}, declared with the
\clj{ns} form, to manage top-level \emph{vars}.
This is a regular Clojure file compiled with
the Clojure compiler except it declares a runtime dependency on
\clj{clojure.core.typed}, Typed Clojure's core namespace.
Programmers import typed versions of certain constructs as needed, like
a \clj{fn} variant that supports annotations,
and use the provided \clj{check-ns} function to type check the current namespace
when convenient.

Typed Clojure can check all the examples in the occurrence typing
paper---the rest of this section describes the extensions necessary
to check Clojure code.

@@

\begin{exmp}
\inputminted[firstline=13,lastline=15]{clojure}{code/demo/src/demo/do.clj}
\end{exmp}

We combine the sequencing form \clj{do} and a thrown exception with
\clj{throw} to avoid a null-pointer exception from attempting to increment \clj{nil}.
(We omit \clj{ns} forms for the rest of the examples)
\footnote{See <this> github repo for full working examples}.

@@

\begin{Code}
\begin{exmp}
\inputminted[firstline=6,lastline=8]{clojure}{code/demo/src/demo/hmap.clj}
\inputminted[firstline=27,lastline=29]{clojure}{code/demo/src/demo/hmap.clj}
\end{exmp}
\end{Code}

@@

HMaps support various options to track the presence and absence of entries.
\clj{:mandatory} declares entries that must be present.

\begin{Code}
\begin{exmp}
\inputminted[firstline=58,lastline=64]{clojure}{code/demo/src/demo/hmap.clj}
\end{exmp}
\end{Code}
Looking up \clj{:a} simply returns the associated type \clj{Num}.
Keys other than \clj{:a} are unaccounted for in \clj{Mand}, so 
the best we can infer is \clj{Any}, the supertype of all types.
\clj{'{:a Num}} is shorthand for \clj{(HMap :mandatory {:a Num})}.

\clj{:absent-keys} is a set of keywords that are known to not have entries.
\begin{Code}
\begin{exmp}
\inputminted[firstline=66,lastline=73]{clojure}{code/demo/src/demo/hmap.clj}
\end{exmp}
\end{Code}
We can infer a better type for \clj{(:b m)}---\clj{:b} is absent in \clj{Abs}
so it looking up \clj{:b} returns \clj{nil}.

\clj{:optional} is a map that declares entries either absent \emph{or}
present with a specific type.

\begin{Code}
\begin{exmp}
\inputminted[firstline=5,lastline=12]{clojure}{code/demo/src/demo/hmap_path.clj}
\end{exmp}
\end{Code}

Concretely \clj{(HMap :optional {:a Num})}
is the same as
\begin{minted}{clojure}
      (U '{:a Num} (HMap :absent-keys #{:a})).
\end{minted}

\clj{:complete?} is \clj{true} if there are no further entries
than those declared \clj{:mandatory} or \clj{:optional}.
\begin{Code}
\begin{exmp}
\inputminted[firstline=17,lastline=26]{clojure}{code/demo/src/demo/hmap_path.clj}
\end{exmp}
\end{Code}
\clj{:a} is \clj{:mandatory} in \clj{Comp} so lookups return \clj{Num}.
\clj{:b} is \clj{:optional} so lookups return \clj{(U nil Kw)}.
\clj{:c} is not mentioned, but since \clj{:complete?} is \clj{true}
it must be absent so a lookup returns \clj{nil}.

@@

HMaps interact with occurrence typing in interesting ways.
The next example branches on the result of a lookup, and the types
inferred for \clj{f} are subtle.

\begin{Code}
\begin{exmp}
\inputminted[firstline=14,lastline=15]{clojure}{code/demo/src/demo/hmap_path.clj}
\end{exmp}
\end{Code}

Typed Clojure infers \clj{m} as \clj{'{:a Num}} in the then branch
and \clj{(HMap :optional {:a nil})}
in the else branch.
We cannot infer 
\begin{minted}{clojure}
 (HMap :absent-keys #{:a})
\end{minted}
in the else branch because \clj{m} could be the value \clj{{:a nil}}
by the definiton of \clj{Opt}.
Section~\ref{sec:hmapformal} discusses this in greater detail.

@@

To resolve reflective calls, Clojure supports \emph{type-hints}.

@@

Typed Clojure ignores type-hints, so to check this we must
annotate the parameter's static type.

@@

Methods are installed via {\clj{defmethod}}.
\begin{minted}{clojure}
(defmethod rep Keyword [x] (str (name x)))
\end{minted}
We extend \clj{rep}'s dispatch table, mapping
the \emph{dispatch value} \clj{Number} to the function
\clj{(fn [x] (str (name x)))}. The call \clj{(rep arg)}
uses the value of \clj{(isa? (disp-fn arg) disp-val)}
on each method entry to pick a winner \clj{mth}, and finally returns \clj{(mth arg)}.
\clj{isa?} is a subclassing check when provided with classes
\begin{minted}{clojure}
  (isa? Keyword Object) ;=> true
  (isa? Keyword Number) ;=> false
\end{minted}
otherwise an equality check.
\begin{minted}{clojure}
  (isa? :a :a) ;=> true
  (isa? :a 1) ;=> false
\end{minted}

@@

\begin{Code}
\begin{exmp}
\inputminted[firstline=5,lastline=11]{clojure}{code/demo/src/demo/rep.clj}
\end{exmp}
\end{Code}
Typed Clojure does not statically ensure multimethod calls dispatch successfully---
\clj{(rep "a")} type checks but throws a runtime error.

The flexibility of \clj{isa?} is key to the generality of multimethods. 
We can dispatch on the \clj{:op} key 
of our HMap AST \clj{Expr}.
Keywords are functions that look themselves up in their argument, so \clj{:op}
is our dispatch function.
\begin{Code}
\begin{exmp}
\inputminted[firstline=5,lastline=23]{clojure}{code/demo/src/demo/eg5.clj}
\end{exmp}
\end{Code}
The destructuring syntax \clj{(fn [{:keys [val] :as m}] ...)} binds
\clj{m} to the value of the first argument, and \clj{val} to \clj{(:val m)}.

@@

The final example combines everything we will cover for the rest of the paper:
multimethod dispatch, reflection resolution via type hints, Java method
and constructor calls, conditional and exceptional flow reasoning
and use of HMaps. 
\begin{Code}
\begin{exmp}
\inputminted[firstline=6,lastline=20]{clojure}{code/demo/src/demo/eg8.clj}
\end{exmp}
\end{Code}
\clj{PayLoad} is either a HMap containing a file
or a string. We dispatch on \clj{:p} to distinguish the two cases---for example on \clj{:F}
we know the \clj{:file} is a file.
The body of the first method uses type-hints to resolve reflection
and conditional control flow to prove null-pointer exceptions are impossible.
The second method is similar except it uses exceptional control flow.

>>>>

We now begin a tour of the central features of Typed Clojure,
beginning with Clojure itself. In our presentation, we will make 
use of the full Typed Clojure system to illustrate the key type system
ideas, before studying the core features in detail in
section~\ref{sec:formal}.

@@

\citet{TF10}
presented Typed Racket with occurrence typing,
a technique for deriving type information from conditional control flow.
They introduce the concept of occurrence typing 
with the following example.

@@

\begin{exmp}
\inputminted[firstline=1]{clojure}{code/demo/src/demo/eg1.clj}
\label{example:conditionalflow}
\end{exmp}

Typed Clojure can check all the examples in~\citet{TF10}---the 
rest of this section describes the extensions necessary
to check Clojure code.

\subsection{Clojure and Typed Clojure Primer}

We introduce Clojure's syntax.
Parenthesis indicate \emph{applications}, square brackets
are syntax for literal \emph{vectors} and curly braces delimit literal
\emph{hash-maps}.
\clj{nil} is exactly Java's \java{null} and is one of two logically
false values (the other is \clj{false}).
\emph{Symbols} begin with an alphabetic character.
A symbol prefixed with a colon like \clj{:a} is an interned \emph{keyword}.
The \clj{get} function takes a hash-map and a key returns the value
associated with the key in the map, or returns \clj{nil} if there is none.
Keywords are functions that take a map and applies \clj{get} to the map
and itself.
\clj{inc} and \clj{dec} increment and decrement numbers respectively.
\clj{assoc} and \clj{dissoc} associate and dissociate entries
from immutable maps.
Everything is an instance of some class
except for \clj{nil}. Predicates like \clj{number?} are equivalent to
a Java \java{instanceof} test, in this case on \clj{java.lang.Number}.

Symbols support \emph{metadata}, an arbitrary map of
data. The syntax \clj{^File x} is a single expression that is the symbol \clj{x} 
with metadata \clj{{:tag File}}.
In binding positions like \clj{(fn [^File x] ...)}
the metadata is preserved in occurrences of \clj{x}.

\emph{Commas} are always whitespace.
\emph{Destructuring} is binding position syntax for pattern matching.
A \emph{map destructuring} contains pairs of symbols and keywords that bind the symbols
to lookups on that keyword with the current argument, optionally terminated by
\clj{:as} and a symbol, which binds the symbol to the current argument.
$$
\clj{(fn [{x :x, y :y, :as m}] ...)}
$$ 
binds \clj{m} to the first argument,
\clj{x} to \clj{(get m :x)} and
\clj{y} to \clj{(get m :y)} in the function body. Type hints are preserved as expected.

Clojure uses \emph{namespaces}, declared with the
\clj{ns} form, to manage top-level \emph{vars} and
class imports.
Example~\ref{example:conditionalflow}
is a regular Clojure file compiled with
the Clojure compiler except it declares a runtime dependency on
\clj{clojure.core.typed}, Typed Clojure's core namespace.
Programmers import typed versions of certain constructs as needed, like
a \clj{fn} variant that supports annotations,
and use the provided \clj{check-ns} function to type check the current namespace
when convenient.
We omit \clj{ns} forms for the rest of the examples.
\footnote{See <this> github repo for full working examples}

Java methods and fields are called in prefix notation
like \clj{(.method target args*)} and \clj{(.field target)} respectively,
with method and field names prefixed with a dot and methods taking some number of arguments.
Constructors are the class name with a dot suffix and arguments like \clj{(class. args*)}.

The \clj{if} special form is a standard two-armed conditional.
The \clj{do} special form sequentially evaluates any number of expressions
left-to-right, returning \nil{} on zero arguments.
The \clj{when} macro is a one-armed \clj{if} that returns \clj{nil} if the test fails.
The \clj{throw} special form is exactly Java's \java{throw}, which takes an
instance of \clj{Throwable} and throws an exception with it.
The \clj{fn} macro creates a function---it takes a parameter vector of symbols indicating parameter names
and any number of expressions in an implicit \clj{do} for the body of the function.
The typed
variant of \clj{fn} supports an optional keyword \clj{:-} and type following
each parameter to declare their static types, and similarly after
the parameter vector to declare the return type.

In a static type, a quoted keyword \clj{':if} is a singleton type containing just the keyword.
Quoted maps \clj{'{:op ':if}} are HMap types---the keys are not quoted because only keyword keys
are allowed. The type constructor \clj{U} indicates an untagged union type, which can take
any number of types.

@@

\begin{exmp}
\inputminted[firstline=13,lastline=15]{clojure}{code/demo/src/demo/do.clj}
\label{example:doexception}
\end{exmp}

@@

\begin{Code}
\begin{exmp}
\inputminted[firstline=6,lastline=13]{clojure}{code/demo/src/demo/hmap.clj}
\end{exmp}
\end{Code}

@@

@@

\begin{Code}
\begin{exmp}
\inputminted[firstline=10,lastline=23]{clojure}{code/demo/src/demo/key2.clj}
\end{exmp}
\end{Code}

HMaps interact with occurrence typing in interesting ways.

\begin{Code}
\begin{exmp}
\inputminted[firstline=15,lastline=27]{clojure}{code/demo/src/demo/hmap.clj}
\end{exmp}
\end{Code}

Section~\ref{sec:hmapformal} discusses this in greater detail.

@@

To resolve reflective calls, Clojure supports \emph{type hints}.

@@

Typed Clojure ignores type hints, so to check this we must
annotate the parameter's static type.

@@

Methods are installed via {\clj{defmethod}}.
\begin{minted}{clojure}
(defmethod rep Keyword [x] (str (name x)))
\end{minted}
We extend \clj{rep}'s dispatch table, mapping
the \emph{dispatch value} \clj{Keyword} to the function
\clj{(fn [x] (str (name x)))}. The call \clj{(rep arg)}
uses the value of \clj{(isa? (disp-fn arg) disp-val)}
on each method entry to pick a winner \clj{mth}, and finally returns \clj{(mth arg)}.
\clj{isa?} is a subclassing check when provided with classes
\begin{minted}{clojure}
  (isa? Keyword Object) ;=> true
  (isa? Keyword Number) ;=> false
\end{minted}
otherwise an equality check.
\begin{minted}{clojure}
  (isa? :a :a) ;=> true
  (isa? :a 1) ;=> false
\end{minted}

@@

\begin{Code}
\begin{exmp}
\inputminted[firstline=5,lastline=11]{clojure}{code/demo/src/demo/rep.clj}
\label{example:rep}
\end{exmp}
\end{Code}
Typed Clojure does not statically ensure multimethod calls dispatch successfully---
\clj{(rep "a")} type checks but throws a runtime error.

The flexibility of \clj{isa?} is key to the generality of multimethods. 
We can dispatch on the \clj{:op} key 
of our HMap AST \clj{Expr}.
Keywords are functions that look themselves up in their argument, so \clj{:op}
is our dispatch function.
\begin{Code}
\begin{exmp}
\inputminted[firstline=5,lastline=25]{clojure}{code/demo/src/demo/eg5.clj}
\end{exmp}
\end{Code}
The destructuring syntax \clj{(fn [{:keys [val] :as m}] ...)} binds
\clj{m} to the value of the first argument, and \clj{val} to \clj{(:val m)}.

@@

The final example combines everything we will cover for the rest of the paper:
multimethod dispatch, reflection resolution via type hints, Java method
and constructor calls, conditional and exceptional flow reasoning
and use of HMaps. 
\begin{Code}
\begin{exmp}
\inputminted[firstline=6,lastline=20]{clojure}{code/demo/src/demo/eg8.clj}
\end{exmp}
\end{Code}
\clj{PayLoad} is either a HMap containing a file
or a string. We dispatch on \clj{:p} to distinguish the two cases---for example on \clj{:F}
we know the \clj{:file} is a file.
The body of the first method uses type hints to resolve reflection
and conditional control flow to prove null-pointer exceptions are impossible.
The second method is similar except it uses exceptional control flow.
